Component response rate variation underlies stability in large complex systems

### 2019-03-02 notebook/SI.Rmd
Revisions, need to add some still to the SI

While I focused on the probability of observing a stable versus unstable system given variation in $\mathbf{\gamma}$ for finite by high system size ($S$), Gibbs et al. [@Gibbs2017] instead focused on the different question concerning the effect of $\mathbf{\gamma}$ on the stability of $\mathbf{M}$ given $\mathbf{A}$ as $S \to \infty$. It is easy to show that my primary results in [the table above](#IncrS) are consistent with Gibbs et al. [@Gibbs2017] by calculating the probability that system stability changes as $S$ increases. Gibbs et al. [@Gibbs2017] predict that the probability that stability is altered should decrease with increasing $S$.

```{r, echo = FALSE}
d_tab    <- dat[,1:7];
p_chnged <- (dat[,6] + dat[,7]) / (dat[,2] + dat[,3]);
gibbtb   <- cbind(d_tab[,1], p_chnged);
colnames(gibbtb) <- c("S", "Pr_changed");
plot(x = gibbtb[,1], y = gibbtb[,2], type = "l");
```

Note, however, that within this finite range of $S$, the probability that an unstable system will be stabilised by $\gamma$ increases as $S$ increases. In other words, the proportion of unstable systems that are stabilised by $\gamma$ increases with $S$ in simulations.

```{r, echo = FALSE}
Pr_stbled <- (dat[,6]) / (dat[,6] + dat[,7]);
gibbtb2   <- cbind(d_tab[,1], Pr_stbled);
colnames(gibbtb2) <- c("S", "Pr_stbled");
plot(x = gibbtb2[,1], y = gibbtb2[,2], type = "l");
```

```

>>>>

I focused on the probability of observing a stable versus unstable system given variation in $\mathbf{\gamma}$ as system complexity ($\sigma\sqrt{SC}$) increased. I increased system complexity by holding $C$ and $\sigma$ constant and incrementally increasing $S$ to obtain numerical results. In contrast, Gibbs et al. [@Gibbs2017] applied analytical techniques to instead focused on a different question concerning the effect of $\mathbf{\gamma}$ on the stability of $\mathbf{M}$ given $\mathbf{A}$ as $S \to \infty$, with $\sigma$ scaled so that $\sigma = 1/\sqrt{S}$. Under such scaling, Gibbs et al. [@Gibbs2017] showed that the effect of $\gamma$ on stability should decrease exponentially as $S$ increases, which I can replicate below by running simulations in which $\sigma = 1/\sqrt{S}$.

The results in the table above can be compared to those of the [main results] (#IncrS). Note that only 100000, rather than 1 million, simulations are run to confirm consistency with Gibbs et al. [@Gibbs2017].

### 2019-03-02 notebook/ms.Rmd
Revisions, need to add some still to the SI

It is important to emphasise that variation in component response rate is not stabilising per se; that is, adding variation in component response rate to a particular system does not necessarily increase the probability that the system will be stable. Rather, highly complex systems that are observed to be stable are more likely to have varying component response rates, and for this variation to be critical to their stability (Fig. 3). This is caused by the shift to a non-uniform distribution of eigenvalues that occurs by introducing $Var(\gamma)$ (Fig. 1b, 2b), which can sometimes cause all of the real components of the eigenvalues of the community matrix $\mathbf{M}$ to become negative, but might also increase the real components of eigenvalues. In this way, my focus is distinct from that of Gibbs et al. [@Gibbs2017], who applied the same mathematical framework to investigate how a diagonal matrix $\mathbf{X}$ (equivalent to $\gamma$ in my model) affects the stability of a community matrix $\mathbf{M}$ given an interaction matrix $\mathbf{A}$ within a generalised Lotka-Volterra model, where $\mathbf{M} = \mathbf{XA}$. Gibbs et al. [@Gibbs2017] analytically demonstrated that the effect of $\mathbf{X}$ on system stability decreases exponentially as system size becomes arbitrarily large ($S \to \infty$). My numerical results are entirely consistent with this prediction (see Supplementary Information). Given that a complex system is known to be stable (or unstable), $Var(\gamma)$ is not expected to increase or decrease the probability of stability. Rather, component response rate variation increases the upper bound at which stability can be realistically observed, meaning that highly complex systems are more likely than not to vary in their component response rates, and for this variation to be critical for system stability.

@@

```{r, eval = TRUE, echo = FALSE, fig.width = 9, fig.height = 7}
dat <- read.csv(file = "sim_results/evolved/evo_results.csv");
dat <- dat[,-1];
plot_stable_4(dat);
```

>>>>

It is important to emphasise that variation in component response rate is not stabilising per se; that is, adding variation in component response rate to a particular system does not necessarily increase the probability that the system will be stable (at high $S$, the opposite is true; see Supplementary Information). Rather, highly complex systems that are observed to be stable are more likely to have varying component response rates, and for this variation to be critical to their stability (Fig. 3). This is caused by the shift to a non-uniform distribution of eigenvalues that occurs by introducing $Var(\gamma)$ (Fig. 1b, 2b), which can sometimes cause all of the real components of the eigenvalues of the community matrix $\mathbf{M}$ to become negative, but might also increase the real components of eigenvalues. In this way, my focus is distinct from that of Gibbs et al. [@Gibbs2017], who applied the same mathematical framework to investigate how a diagonal matrix $\mathbf{X}$ (equivalent to $\gamma$ in my model) affects the stability of a community matrix $\mathbf{M}$ given an interaction matrix $\mathbf{A}$ within a generalised Lotka-Volterra model, where $\mathbf{M} = \mathbf{XA}$. Gibbs et al. [@Gibbs2017] analytically demonstrated that the effect of $\mathbf{X}$ on system stability decreases exponentially as system size becomes arbitrarily large ($S \to \infty$). My numerical results are not inconsistent with this prediction because I did not scale $\sigma = 1 / \sqrt{S}$, but instead fixed $\sigma$ and increased $S$ to thereby increase total system complexity, $\sigma\sqrt{SC}$ (see Supplemental Information for results simulated across $\sigma$ and $C$. Overall, my results show that component response rate variation increases the upper bound of complexity at which stability can be realistically observed, meaning that highly complex systems are more likely than not to vary in their component response rates, and for this variation to be critical for system stability.

@@

### 2019-03-02 notebook/ms.tex
Revisions, need to add some still to the SI

\subsection{Author Summary}\label{author-summary}

Any set of discrete components that potentially interact can define a
complex system. Typically, properties of individual components will
cause them to respond to system perturbation at different rates, but the
consequences of such variation for system stability are unexplored. This
work makes three novel theoretical contributions: First, I show that
large complex systems with varying component response rates are more
likely to be stable than when response rates are uniform. Second, I show
that a targeted manipulation of component response rates can increase
the probability of system stability by several orders of magnitude.
Lastly, I show that varying component response rate does not affect
system feasibility. These results are general to any physical,
biological, social, or mixed system.

@@

It is important to emphasise that variation in component response rate
is not stabilising per se; that is, adding variation in component
response rate to a particular system does not necessarily increase the
probability that the system will be stable. Rather, highly complex
systems that are observed to be stable are more likely to have varying
component response rates, and for this variation to be critical to their
stability (Fig. 3). This is caused by the shift to a non-uniform
distribution of eigenvalues that occurs by introducing \(Var(\gamma)\)
(Fig. 1b, 2b), which can sometimes cause all of the real components of
the eigenvalues of the community matrix \(\mathbf{M}\) to become
negative, but might also increase the real components of eigenvalues. In
this way, my focus is distinct from that of Gibbs et
al.\textsuperscript{\protect\hyperlink{ref-Gibbs2017}{17}}, who applied
the same mathematical framework to investigate how a diagonal matrix
\(\mathbf{X}\) (equivalent to \(\gamma\) in my model) affects the
stability of a community matrix \(\mathbf{M}\) given an interaction
matrix \(\mathbf{A}\) within a generalised Lotka-Volterra model, where
\(\mathbf{M} = \mathbf{XA}\). Gibbs et
al.\textsuperscript{\protect\hyperlink{ref-Gibbs2017}{17}} analytically
demonstrated that the effect of \(\mathbf{X}\) on system stability
decreases exponentially as system size becomes arbitrarily large
(\(S \to \infty\)). My numerical results are entirely consistent with
this prediction (see Supplementary Information). Given that a complex
system is known to be stable (or unstable), \(Var(\gamma)\) is not
expected to increase or decrease the probability of stability. Rather,
component response rate variation increases the upper bound at which
stability can be realistically observed, meaning that highly complex
systems are more likely than not to vary in their component response
rates, and for this variation to be critical for system stability.

@@

\textbf{Figure 2: Distributions of eigenvalues before (a) and after (b)
introducing variation in component response rate
(\(\boldsymbol{\gamma}\)) in complex systems.} Each panel show the same
system where \(S = 1000\), \(C = 0.05\), and \(\sigma = 0.4\).
\textbf{a.} Eigenvalues plotted in the absence of \(Var(\gamma)\) where
\(E[\gamma] = 1\), versus \textbf{b.} eigenvalues plotted given
\(\gamma \sim \mathcal{U}(0, 2)\), which increases the variance of
interaction strengths (\(\sigma^{2}\)) but also creates a cluster of
eigenvalues toward the distribution's centre (-1, 0). Blue elipses in
both panels show the circle centred on the distribution in panel a.
Proportions of \(\Re(\lambda) < 0\) are 0.724 and 0.738 for a and b,
respectively.

@@

\includegraphics{ms_files/figure-latex/unnamed-chunk-16-1.pdf}

>>>>

@@

It is important to emphasise that variation in component response rate
is not stabilising per se; that is, adding variation in component
response rate to a particular system does not necessarily increase the
probability that the system will be stable (at high \(S\), the opposite
is true; see Supplementary Information). Rather, highly complex systems
that are observed to be stable are more likely to have varying component
response rates, and for this variation to be critical to their stability
(Fig. 3). This is caused by the shift to a non-uniform distribution of
eigenvalues that occurs by introducing \(Var(\gamma)\) (Fig. 1b, 2b),
which can sometimes cause all of the real components of the eigenvalues
of the community matrix \(\mathbf{M}\) to become negative, but might
also increase the real components of eigenvalues. In this way, my focus
is distinct from that of Gibbs et
al.\textsuperscript{\protect\hyperlink{ref-Gibbs2017}{17}}, who applied
the same mathematical framework to investigate how a diagonal matrix
\(\mathbf{X}\) (equivalent to \(\gamma\) in my model) affects the
stability of a community matrix \(\mathbf{M}\) given an interaction
matrix \(\mathbf{A}\) within a generalised Lotka-Volterra model, where
\(\mathbf{M} = \mathbf{XA}\). Gibbs et
al.\textsuperscript{\protect\hyperlink{ref-Gibbs2017}{17}} analytically
demonstrated that the effect of \(\mathbf{X}\) on system stability
decreases exponentially as system size becomes arbitrarily large
(\(S \to \infty\)). My numerical results are not inconsistent with this
prediction because I did not scale \(\sigma = 1 / \sqrt{S}\), but
instead fixed \(\sigma\) and increased \(S\) to thereby increase total
system complexity, \(\sigma\sqrt{SC}\) (see Supplemental Information for
results simulated across \(\sigma\) and \(C\). Overall, my results show
that component response rate variation increases the upper bound of
complexity at which stability can be realistically observed, meaning
that highly complex systems are more likely than not to vary in their
component response rates, and for this variation to be critical for
system stability.

@@

\textbf{Figure 2: Distributions of eigenvalues before (a) and after (b)
introducing variation in component response rate
(\(\boldsymbol{\gamma}\)) in complex systems.} Each panel show the same
system where \(S = 1000\), \(C = 0.05\), and \(\sigma = 0.4\).
\textbf{a.} Eigenvalues plotted in the absence of \(Var(\gamma)\) where
\(E[\gamma] = 1\), versus \textbf{b.} eigenvalues plotted given
\(\gamma \sim \mathcal{U}(0, 2)\), which increases the variance of
interaction strengths (\(\sigma^{2}\)) but also creates a cluster of
eigenvalues toward the distribution's centre (-1, 0). Blue elipses in
both panels show the circle centred on the distribution in panel a.
Proportions of \(\Re(\lambda) < 0\) are 0.726 and 0.742 for a and b,
respectively.

@@
